A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-05 paper.tex
Draft of first half of disco

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
meaning that sample size is substantially larger than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination,
but for which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

>>>>

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.
A necessary first step is a degree of  terminological clarity.
As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture complexities of family structure which
make, for example, polytomies a necessary feature of local trees.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
were a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner such an  incompehensibly
large space of possibilities.
It is therefore important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here,
is to use polytomies and stacked recombinations as direct topological
descriptions of uncertainty. Other methods to capture, for example,
uncertainty about node ages and breakpoint location, is an important
aspect of future work.

@@

Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination,
but for which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.
